# Enrichment Analysis

```{r Setup, include=FALSE, message=FALSE}
library(dplyr)
library(ggplot2)
library(magrittr)
library(readr)
library(reshape2)
library(knitr)
library(tibble)

# Directory Setup
dataDir <- file.path(getwd(), "data")
enrichmentDir <- file.path(dataDir, "enrichment")

importHomerFuncEnr <- function(path, p.adjust.method = "fdr", a = 0.05, 
                                whichEnr = c("biological_process.txt", 
                                                  "molecular_function.txt", 
                                                  "cellular_component.txt", 
                                                  "msigdb.txt", 
                                                  "reactome.txt", 
                                                  "interpro.txt", 
                                                  "kegg.txt", 
                                                  "prosite.txt")) {
  
  # Get the paths to the enrichment analysis results. 
  enrFilePaths <- paste0(path, whichEnr) %>%
    set_names(gsub(x = whichEnr, pattern = ".txt", replacement = ""))
  
  # Import the enrichment analysis results, perform multiple testing p-value adjustment, 
  # and then perform filtering for significant p-values. 
  importedEnr <- lapply(enrFilePaths, function(x) readr::read_tsv(x, col_types = "?????????c?")) 
  importedEnr <- do.call("rbind", importedEnr) %>%
    rownames_to_column("Source") %>%
    mutate(Source = gsub(x = Source, pattern = "\\.[[:digit:]]{1,3}$", replacement = "")) %>%
    mutate(AdjustedP = p.adjust(Enrichment, p.adjust.method)) %>%
    filter(AdjustedP < a)
  
  return(importedEnr)
}

bonAdjust <- function(x, a = 1) {
  read_tsv(x) %>%
    dplyr::select(-`P-value`, -`q-value (Benjamini)`) %>%
  mutate(PValue = exp(`Log P-value`),
         Bonferroni = p.adjust(PValue, "bonferroni"),
         FDR = p.adjust(PValue, "fdr")) %>%
  filter(PValue < a)
}
```


## Define Groups

In this section, we define groups of genes that will be tested with the following:

- **Functional Enrichment**: Gene Ontology, KEGG, Interpro, MSigDB gene sets, etc. using HOMER software. 
- **Promoter Motif Enrichment**: Tested for known zebrafish gene promoter motifs corresponding to transcription factor binding sites, using HOMER software. 

First, we will import the results of the DE gene analysis in the `limma_results` data.frame and use this to define the gene groups. 

```{r Define-Gene-Groups, message=FALSE, warning=FALSE}
# Import results from DE analysis 
limma_results <- readRDS(file.path(dataDir, "de_genes", "limma_results.rds"))

# Export the list of genes in limma_results for use as the "background" set of genes for enrichment testing. 
# limma_results %>% dplyr::select(ensembl_gene_id) %>% write_tsv(file.path(enrichmentDir, "backgroundGenes.txt"))


# Define the groups of genes depending on whether they are up (1) or down (-1) regulated in particular comparisons. 
geneGroups <- list(
  inappropriatelyDownregulated = limma_results %>% filter(Res.K97_mutOld_wtOld == -1, Res.K97_mutOld_mutYoung == -1),
  failureToUpregulate = limma_results %>% filter(Res.K97_wtOld_wtYoung == 1, Res.K97_mutOld_wtOld == -1),
  failureToDownregulate = limma_results %>% filter(Res.K97_wtOld_wtYoung == -1, Res.K97_mutOld_wtOld == 1),
  acceleratedAging = limma_results %>% filter((Res.K97_mutYoung_wtYoung == 1 & Coef.K97_wtOld_wtYoung > 0 & p.value.K97_wtOld_wtYoung < 0.05) |
                                               (Res.K97_mutYoung_wtYoung == -1 & Coef.K97_wtOld_wtYoung < 0 & p.value.K97_wtOld_wtYoung < 0.05)),
  agingSignature = limma_results %>% filter((Res.K97_wtOld_wtYoung == -1 & Res.K97_mutOld_mutYoung == -1) | 
                                              (Res.K97_wtOld_wtYoung == 1 & Res.K97_mutOld_mutYoung == 1)) 
  
)

# Define groups of DE genes for each comparison.
geneGroups_DE <- list(
  mutYoungwtYoung = limma_results %>% filter(Res.K97_mutYoung_wtYoung != 0),
  mutOldwtOld = limma_results %>% filter(Res.K97_mutOld_wtOld != 0),
  wtOldwtYoung = limma_results %>% filter(Res.K97_wtOld_wtYoung != 0),
  mutOldmutYoung = limma_results %>% filter(Res.K97_mutOld_mutYoung != 0)
)

# Display the number of genes in each group. 
geneGroups %>% lapply(dim) %>% as.data.frame %>% extract(1,) %>% kable(caption = "Number of genes in each group.")
geneGroups_DE %>%  lapply(dim) %>% as.data.frame %>% extract(1,) %>% kable(caption = "Number of DE genes for each comparison.")
```

## Motif Enrichment with HOMER

*HOMER* was run using the [findMotifs.pl](http://homer.ucsd.edu/homer/motif/index.html) method to obtain the enriched known zebrafish promoter motifs and functional enrichment. Note that the background set of genes used when running *HOMER* is the `r dim(limma_results)[1]` genes in the `limma_results` object from the DE gene analysis. 

Below is an example of how *HOMER*'s `findMotifs.pl` method can be run using a list of gene names (ENSEMBL format in this case), a list of curated zebrafish promoters

```{r HOMER-Motif-Enrichment-Example, engine='bash', eval=FALSE}
findMotifs.pl inappropriatelyDownregulated.txt \ #Input list of genes. 
zebrafish \ #Use known zebrafish promoter motifs to look for enrichment. 
inappropriatelyDownregulated \ #Folder to save the results into. 
-start -1500 -end 200 \ #Define promoter region as 1500bp upstream and 200bp downstream.
-p 4 \ #Use four CPUs
-bg backgroundGenes.txt #List of background genes. 
```

## Functional Enrichment Results

We can now import the results using `importHomerFuncEnr` which merges together the enrichment results from each set of tests (e.g. Gene ontology, KEGG, Interpro, etc. ) and does multiple testing adjustment (Bonferroni) across the whole merged data.frame. 

```{r Functional-Enrichment-Results, error=FALSE, warning=FALSE, message=FALSE}
# Gene Groups
inappropriatelyDownregulated <- importHomerFuncEnr(file.path(enrichmentDir, "inappropriatelyDownregulated/"), "bonferroni")
failureToUpregulate <- importHomerFuncEnr(file.path(enrichmentDir, "failureToUpregulate/"), "bonferroni")
failureToDownregulate <- importHomerFuncEnr(file.path(enrichmentDir, "failureToDownregulate/"), "bonferroni")
acceleratedAging <- importHomerFuncEnr(file.path(enrichmentDir, "acceleratedAging/"), "bonferroni")
agingSignature <- importHomerFuncEnr(file.path(enrichmentDir, "commonAgingGenes/"), "bonferroni")
inversion <- importHomerFuncEnr(file.path(enrichmentDir, "inversion/"), "bonferroni")

# Gene groups from DE analysis
mutYoungwtYoung <- importHomerFuncEnr(file.path(enrichmentDir, "mutYoungwtYoung/"), "bonferroni")
mutOldwtOld <- importHomerFuncEnr(file.path(enrichmentDir, "mutOldwtOld/"), "bonferroni")
wtOldwtYoung <- importHomerFuncEnr(file.path(enrichmentDir, "wtOldwtYoung/"), "bonferroni")
mutOldmutYoung <- importHomerFuncEnr(file.path(enrichmentDir, "mutOldmutYoung/"), "bonferroni")



functionalEnrichmentResults <- list(inappropriatelyDownregulated = inappropriatelyDownregulated, 
                                    failureToUpregulate = failureToUpregulate, 
                                    failureToDownregulate = failureToDownregulate, 
                                    acceleratedAging = acceleratedAging,
                                    agingSignature = agingSignature,
                                    inversion = inversion,
                                    mutYoung_vs_wtYoung = mutYoungwtYoung,
                                    mutOld_vs_wtOld = mutOldwtOld,
                                    wtOld_vs_wtYoung = wtOldwtYoung,
                                    mutOld_vs_mutYoung = mutOldmutYoung
                                    )

# Export the functional enrichment results. 
# writexl::write_xlsx(functionalEnrichmentResults, file.path(enrichmentDir, "functionalEnrichmentResults.xlsx"))


# Preview the Results
functionalEnrichmentResults %>% lapply(., function(x) { x %>% as.data.frame %>% head %>% kable})
```

## Promoter Motif Enrichment results

*HOMER* outputs the promoter enrichment results in the `knownResults.txt` table for each set of genes, which contains the unadjusted enrichment *p*-values for each promoter motif. Here, we will import this table and perform Bonferroni adjustment on the *p*-values. 

```{r Motif-Enrichment-Results, error=FALSE, warning=FALSE, message=FALSE}
motifs <- list(
  inappropriatelyDownregulated = bonAdjust(file.path(enrichmentDir, "inappropriatelyDownregulated", "knownResults.txt"), 1),
  failureToUpregulate = bonAdjust(file.path(enrichmentDir, "failureToUpregulate", "knownResults.txt"), 1),
  failureToDownregulate = bonAdjust(file.path(enrichmentDir, "failureToDownregulate", "knownResults.txt"), 1),
  acceleratedAging = bonAdjust(file.path(enrichmentDir, "acceleratedAging", "knownResults.txt"), 1),
  inversion = bonAdjust(file.path(enrichmentDir, "inversion", "knownResults.txt"), 1),
  mutYoungwtYoung = bonAdjust(file.path(enrichmentDir, "mutYoungwtYoung", "knownResults.txt"), 0.05),
  mutOldwtOld = bonAdjust(file.path(enrichmentDir, "mutOldwtOld", "knownResults.txt"), 0.05)
)

# Export the functional enrichment results. 
# writexl::write_xlsx(motifs, file.path(enrichmentDir, "motifEnrichmentResults.xlsx"))

# Preview the results:
motifs %>% lapply(., function(x) { x %>% as.data.frame %>% head %>% kable})
```

